Performance Evaluation of Neural Networks for Intrusion Prevention Systems

### 2020-06-30 paper2/paper2.tex
Added abstract and implemented feedback from Boris

\title{Title}

@@

\begin{abstract}
@TBA
\end{abstract}

\begin{IEEEkeywords}
artificial intelligence, neural networks, computer networks, network security
\end{IEEEkeywords}

@@

It is suspected that there are differences in the intrusion detection datasets because, for example, cyber attacks are not always executed or recorded in the same way or the network infrastructure is structured differently. The aim is to determine how sensitive neural networks react to this and are thus dependent on the data with which they have been trained. This chapter is designed in such a way that the various steps for building up the neural network are first described and then the results obtained are examined in detail.

@@

\subsection{Training the neural network}
After the preprocessing of the data has been completed, the next task is to train and test the neural network. The architecture recommended in \cite{max1} (three hidden layers with 69 neurons each) and the hyperparameters (Learn rate: 0. 0001, Optimizer: adam, Weight initialization: normal, Activation function: relu, Number of neurons per layer: 69, Epochs: 850, Batch size: 256) are reused. In addition, "MinMaxScale" was used to normalize the data and a weight value of 500 was set for benign" traffic. At this point it is worth mentioning that attempts have been made to adjust individual parameters such as the learn rate or the weight value of benign traffic in order to achieve a better result. However, this has not had any positive effects. Future work could deal with this in more detail and try to find optimal hyperparameters by an extensive gridsearch. Of course, this would require appropriate hardware resources.

@@

\begin{figure}[htbp]  
\centerline{\includegraphics[scale=0.65]{NeuesModellNeueDaten.png}}
\caption{Confusion Matrix for Szenario 1}
\label{fig:cm1}
\end{figure}

@@

\begin{figure}[htbp]  
\centerline{\includegraphics[scale=0.65]{NeuesModellAlteDaten.png}}
\caption{Confusion Matrix for Szenario 2}
\label{fig:cm2}
\end{figure}

In the third scenario, the neural network from \cite{max1}, which was trained with CSE-CIC-IDS2018 Dataset \cite{max2}, is now tested with the dataset of this work. This continues the low classification success (see Figure \ref{fig:cm3}). In the class "benign", 99.8\% of all data are correctly predicted (recall), but the precision is noticeable low (0. 008\%). This means that the neural network recognizes many data points as "benign", but these are mostly other classes. Conversely, this is the case with "bot", where only few data of this class are correctly recognized, but whenever the neural network recognizes this class, it is a correct prediction. All other precision and recall values are or move around 0\%. It is particularly noticeable that "bot"-, "ssh-bruteforce"- and "dos-goldeneye" attacks are mostly classified as benign traffic. Furthermore, it can be seen that "ssh-bruteforce"- and "dos-goldeneye" attacks cannot be correctly detected once; bot attacks can be correctly predicted nine times in total. This makes it clear how weak the classifications of the neural network trained in \cite{max1} are with the dataset created in this work. Also worth mentioning is that for some data points the classes "ddos attacks-loic-http", "dos attacks-hulk" and "dos attacks-slowloris" are predicted. These are not available in the dataset created in this work, but in the CSE-CIC-IDS2018 dataset \cite{max2}, with which the neural network was trained. For this reason, this is not surprising.

\begin{figure}[htbp]  
\centerline{\includegraphics[scale=0.65]{AltesModellNeueDaten.png}}
\caption{Confusion Matrix for Szenario 3}
\label{fig:cm3}
\end{figure}

@@

\section{Conclusion}\label{conclusion}
In summary, the performance of neural networks in terms of intrusion detection over different datasets is rather mixed. If these machine learning algorithms are tested with the same dataset with which they were trained, the result is promising. Using another dataset for testing results in a poor classification performance. However, there is lots of potential for future research with neuronal networks in terms of intrusion detection. It seems useful to analyse in detail why the mentioned difficulties occur in the classification and which factors are relevant for the neural network. An intensive examination of the finding of optimal hyperparameters would also be desirable in the course of future work. Furthermore, it would also make sense to simplify the way classification is done. Instead, a simple classification like network traffic is “benign” or “not benign” would be conceivable. Moreover, it should be checked whether a better performance can be achieved with other machine learning algorithms.

>>>>

\title{Performance Evaluation of Neural Networks for Intrusion Prevention Systems}

@@

\begin{abstract}
The detection of malicious behavior in network traffic through the use of machine learning algorithms in intrusion prevention systems is still a current topic of research. This paper focuses on neural networks and examines their performance across different datasets. Basically, the quality of the data with which machine learning algorithms are trained is a decisive factor for the classification performance. In addition to the publicly available CSE-CIC-IDS2018 dataset, this work also uses a dataset that was created with the help of feature extraction from automatically generated network traffic. Different scenarios are tested, but it turns out that the classification performance of neural networks in this context is only satisfactory if they are tested with data from the dataset with which they were trained. If another dataset is used for testing, the result of the classification is poor. This indicates that neural networks react sensitively to differences in the datasets and that the classification performance is influenced by this.
\end{abstract}

\begin{IEEEkeywords}
artificial intelligence, neural networks, computer networks, network security, feature extraction
\end{IEEEkeywords}

@@

It is expectable that there are differences in the intrusion detection datasets because, for example, cyber attacks are not always executed or recorded in the same way or the network infrastructure is structured differently. The aim is to determine how sensitive neural networks react to this and are thus dependent on the data with which they have been trained. This chapter is designed in such a way that the various steps for building up the neural network are first described and then the results obtained are examined in detail.

@@

\subsection{Training the neural network}
After the preprocessing of the data has been completed, the next task is to train and test the neural network. The architecture recommended in \cite{max1} (three hidden layers with 69 neurons each) and the hyperparameters (Learn rate: 0. 0001, Optimizer: adam, Weight initialization: normal, Activation function: relu, Number of neurons per layer: 69, Epochs: 850, Batch size: 256) are reused. In addition, "MinMaxScale" was used to normalize the data and a weight value of 500 was set for benign traffic. At this point it is worth mentioning that attempts have been made to adjust individual parameters such as the learn rate or the weight value of benign traffic in order to achieve a better result. However, this has not had any positive effects. Future work could deal with this in more detail and try to find optimal hyperparameters by an extensive gridsearch. Of course, this would require appropriate hardware resources.

@@

\begin{figure}[htbp]  
\centerline{\includegraphics[scale=0.65]{NeuesModellNeueDaten.png}}
\caption{Confusion Matrix for Scenario 1}
\label{fig:cm1}
\end{figure}

@@

\begin{figure}[htbp]  
\centerline{\includegraphics[scale=0.65]{NeuesModellAlteDaten.png}}
\caption{Confusion Matrix for Scenario 2}
\label{fig:cm2}
\end{figure}

In the third scenario, the neural network from \cite{max1}, which was trained with CSE-CIC-IDS2018 Dataset \cite{max2}, is now tested with the dataset of this work. This continues the low classification success (see Figure \ref{fig:cm3}). In the class "benign", 99.8\% of all data are correctly predicted (recall), but the precision is noticeably low (0. 008\%). This means that the neural network recognizes many data points as "benign", but these are mostly other classes. Conversely, this is the case with "bot", where only few data of this class are correctly recognized, but whenever the neural network recognizes this class, it is a correct prediction. All other precision and recall values are or move around 0\%. It is particularly noticeable that "bot"-, "ssh-bruteforce"- and "dos-goldeneye" attacks are mostly classified as benign traffic. Furthermore, it can be seen that "ssh-bruteforce"- and "dos-goldeneye" attacks cannot be correctly detected once; bot attacks can be correctly predicted nine times in total. This makes it clear how weak the classifications of the neural network trained in \cite{max1} are with the dataset created in this work. Also worth mentioning is that for some data points the classes "ddos attacks-loic-http", "dos attacks-hulk" and "dos attacks-slowloris" are predicted. These are not available in the dataset created in this work, but in the CSE-CIC-IDS2018 dataset \cite{max2}, with which the neural network was trained. For this reason, this is not surprising.

\begin{figure}[htbp]  
\centerline{\includegraphics[scale=0.65]{AltesModellNeueDaten.png}}
\caption{Confusion Matrix for Scenario 3}
\label{fig:cm3}
\end{figure}

@@

\section{Conclusion}\label{conclusion}
In summary, the performance of neural networks in terms of intrusion detection over different datasets is rather mixed. If these machine learning algorithms are tested with the same dataset with which they were trained, the result is promising. Using another dataset for testing results in a poor classification performance. However, there is lots of potential for future research with neural networks in terms of intrusion detection. It seems useful to analyse in detail why the mentioned difficulties occur in the classification and which factors are relevant for the neural network. An intensive examination of the finding of optimal hyperparameters would also be desirable in the course of future work. Furthermore, it would also make sense to simplify the way classification is done. Instead, a simple classification like network traffic is “benign” or “not benign” would be conceivable. Moreover, it should be checked whether a better performance can be achieved with other machine learning algorithms.
